
```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
```


## Exercises

The exercises revisit our `grades` data:

```{r echo = FALSE}
# Get rid of some duplicate rows!
grades <- read.csv("https://mac-stat.github.io/data/grades.csv") |> 
  distinct(sid, sessionID, .keep_all = TRUE)

# Check it out
head(grades)
```

We'll explore the number of times each grade was assigned:

```{r}
grade_distribution <- grades |> 
  count(grade)

head(grade_distribution)
```

### Exercise 1: Changing the order (option 1) {.unnumbered}

Check out a **column plot** of the number of times each grade was assigned during the study period. This is similar to a bar plot, but where we define the height of a bar according to variable in our dataset.

```{r}
grade_distribution |> 
  ggplot(aes(x = grade, y = n)) +
    geom_col()
```

The order of the grades is goofy! Construct a new column plot, manually reordering the grades from high (A) to low (NC) with "S" and "AU" at the end:

```{r}
 grade_distribution |>
   mutate(grade = fct_relevel(grade, c("A", "A-", "B+", "B", "B-", "C+", "C", "C-", "D+", "D", "D-", "NC", "S", "AU"))) |>
   ggplot(aes(x = grade, y = n)) +
     geom_col()
```

Construct a new column plot, reordering the grades in ascending frequency (i.e. how often the grades were assigned):

```{r}
 grade_distribution |>
   mutate(grade = fct_reorder(grade, n)) |>
   ggplot(aes(x = grade, y = n)) +
    geom_col()
```

Construct a new column plot, reordering the grades in descending frequency (i.e. how often the grades were assigned):

```{r}
 grade_distribution |>
   mutate(grade = fct_reorder(grade, n, .desc = TRUE)) |>
   ggplot(aes(x = grade, y = n)) +
     geom_col()
```

\
\
\
\

### Exercise 2: Changing factor level labels {.unnumbered}

It may not be clear what "AU" and "S" stand for. Construct a new column plot that renames these levels "Audit" and "Satisfactory", while keeping the other grade labels the same *and* in a meaningful order:

```{r}
 grade_distribution |>
   mutate(grade = fct_relevel(grade, c("A", "A-", "B+", "B", "B-", "C+", "C", "C-", "D+", "D", "D-", "NC", "S", "AU"))) |>
   mutate(grade = fct_recode(grade, "Audit" = "AU", "Satisfactory" = "S")) |>  # Multiple pieces go into the last 2 blanks
   ggplot(aes(x = grade, y = n)) +
     geom_col()
```
